---
title: "Spotify Playlist Manager- Sta 323 Final Project"
author: "Garrett Allen"
output:
  pdf_document: default
  html_document: default
---

## Setup

```{r setup, include = FALSE}
library(tidyr)
library(tidyverse)
library(spotifyr)
library(glue)
library(shiny)
library(DT)
library(devtools)
devtools::install_github("EdwinTh/ggoutlier")
library(ggoutlier)
library(shinyalert)
```

## Introduction

```{r getting auth code, eval = FALSE}
#setting up API. Make sure when get_spotify_authorization_code runs, you 
#verify with spotify and select "No" when it asks if you'd like to cache results in Console. 
#RUN THIS BEFORE ALL OTHER CODE

obtain id/secret from setting up a spotify app using the spotify api. 

id <- put spotify id here
secret <- put secret id here

Sys.setenv(SPOTIFY_CLIENT_ID = id)
Sys.setenv(SPOTIFY_CLIENT_SECRET = secret)
access_token <- get_spotify_access_token()
auth_code <- get_spotify_authorization_code()
```
For this project, I wanted to make something really practical that I would use. 
After I thought for awhile, I realized a common issue I have is that there's 
not a lot of nice tools for Spotify to manage playlists. I store all my music 
in one large mega-playlist (about ~3000 songs), which, while I like, doesn't 
let me easily analyze my music or communicate my taste in music to others. 
For example, if I wanted to send all of my music of a certain genre, or a 
certain feel, I'd have no way of doing so easily without manually doing so. 
This is clearly intractable for me now with the amount of music in my playlist,
and it's even worse for some of my friends, who have playlists with > 10,000 songs.

As a result, I resolved that I wanted to make a ShinyApp that made it easy to do 
a few key tasks with playlists. For one, I wanted to be able to filter my 
playlists by a variety of features easily, as well as have visualizations that
let me understand both the distribution of my music and how to effectively filter 
it. I also wanted to make the app capable of handling Liked Songs, an internal 
playlist that Spotify doesn't normally let you turn into public playlists. My 
friend has all of his music in his Liked Songs, and it makes it really difficult
for him to share music in a mass-way. 

Finally, I wanted to have a clean implementation to combine playlists together,
as the current method of dragging and dropping in the Spotify client becomes
really cumbersome for large datasets. 

## Methods / Implementation

To implement all of this, I used a variety of packages. SpotifyR is a nice 
wrapper around the Spotify API so that I don't have to deal with some of deeper
intricacies of the Spotify API. While I considered foregoing the package at 
first so I could have more flexiility, I found that the wrapper had more than 
enough flexibility to be useful for my purposes. 

I also used a small package called ggoutlier to make a nice histogram 
that avoided being too skewed due to high follower counts of a few artists. 
The rest of the packages are fairly standard or self explanatory. 

One of the most annoying aspects of this project was figuring out how to use an
authorization code with Spotify's OAuth 2.0 system. This is necessary to access
data from Spotify users that is not public, as well as necessary to create
playlists from song ids. The details are not so important now, but the 
documentation on Spotify's website is rather poor.

It may be necessary for me to grant you access to my Spotify app in development
mode for the OAuth to work properly; if that is true, please email me so I can
add both your email and name registered to your Spotify account to Spotify's API
manager. It should all work and be reproducible, as I tested it both on my 
computer and a friend's, but getting the browser to properly authenicate with 
R has been a bit of a difficulty, so fingers crossed. 

After getting the API keys from Spotify and implementing OAuth 2.0 in my 
Spotify App Developer page, I decided that I wanted to make a few functions that
would make my life easier (specifically around Pagination) and then leverage
these functions in my app to make the code cleaner. The pagination functions
are all rather similar, although there had to be minor adjustments based on the
specific operation happening. They are all variations of grabbing x results at
a time and then shifting upwards by x in position to grab the next x results. 

The main meat of the project is the ShinyApp itself; I did end up implementing 
every aspect that I wanted, and I am now able to filter Spotify Playlists/Liked
Songs by tempo, danceability, energy, valence (happiness), speechiness, 
acousticness, number of followers the artist has, and the genre, as well as
combine playlists together and extract these playlists to my Spotify account. 

A few notes on using the app: to get a playlist_id, one needs to first 
create a playlist on Spotify, then get the share link. After getting the link to
share it, you must extract the part between playlist/ and ?. I have loaded 
some default values for ease of verifying that the app works.

For example, in the link below, the ID would be 5SNDR83wfKeMvWhcablylH.

https://open.spotify.com/playlist/5SNDR83wfKeMvWhcablylH?si=c5f38043f81b4114

The app is rather robust to failed inputs due to a lot of tryCatch loops 
that validate various sections of code. It should not crash easily.

Spotify's requests per day is very generous, and I never ran into an issue 
where Spotify ratelimited my usage of their API, so I have not put in to place
any safeguards to protect against ratelimiting; I don't anticipate this will be 
an issue, as I made upwards of hundreds of calls in a short amount of time 
without triggering any ratelimiting. 

I used a navPage to make the app easier to navigate, along with several 
tabPanels that made the main panel significantly less cluttered. In order
to generate the data seen in the main panel under "Songs", I had to extract
songs from a playlist using get_all_playlist_tracks, then extract more information
from the tracks with get_all_x_features using the function get_track_audio_features,
and then pull artist information from get_all_x_features with the function
get_artists. Note that tracks with more than 1 artists are duplicated
in the output, as Spotify classes the genre of a song by the artists' genre. 
That is, Spotify does no track level genre classification, so its important
to look at the genre of all artists collaborating on a track in order to get
a clear view of the genres present. I then joined all the dataframes together
with joining functions, using various unique IDS to combine them together. 

After getting the dataset, the code is mostly straightforward; it applies
filtering if the user selects it, generates a playlist if the user selects it
using the methods create_playlist and add_all_tracks_to_playlist, and generates
a few visualizations. Updates only occur upon pressing the action button at the 
bottom, as we would hope. 

If you try a large playlist with the app, give it some time to generate results:
pagination makes some of the operations quite slow, particularly the artist one,
as results can only be grabbed 20 at a time. Results are still fairly fast on
playlists of thousands, however.

```{r pagination functions}

#gets all songs in a playlist
get_all_playlist_tracks <- function(id){
  my_tracks <- get_playlist_tracks(playlist_id = id)
  new_page = 0
  prev_tracks = 1
  #handles pagination for indeterminate size of playlist by going until my_tracks doesn't increase in size
  while(nrow(my_tracks) != prev_tracks){ 
    prev_tracks = nrow(my_tracks)
    new_page = new_page + 100
    new_tracks <- get_playlist_tracks(playlist_id = id, offset = new_page)
    my_tracks <- rbind(my_tracks,new_tracks)
    print(glue("get_all_playlist_tracks: Processing page {new_page / 100} of playlist"))
  }
  my_tracks %>% #filtering bad row
    select(-(2:4),-(6:8),-13,-17,-21,-23,-33,-40)
  
}

#gets saved tracks from user's library, similar to previous method
get_all_saved_tracks <- function(auth_code){ 
  my_tracks <- get_my_saved_tracks(limit = 20)
  new_page = 0
  prev_tracks = 1
  while(nrow(my_tracks) != prev_tracks){
    prev_tracks = nrow(my_tracks)
    new_page = new_page + 20
    new_tracks <- get_my_saved_tracks(offset = new_page, authorization = auth_code)
    my_tracks <- rbind(my_tracks,new_tracks)
    print(glue("Processing page {new_page / 20}"))
  }
  my_tracks
  
}

#gets all "features" from a type of thing. written generically to allow
# for getting album features, track features, artist features, such as 
#genre, followers, internal spotify data, etc. handles pagination 
#enforced by spotifyR package. 
get_all_x_features <- function(all_ids,limit,func){ 
  
  if(length(all_ids) <= limit){
    return(func(ids = all_ids))
  }
  
  features <- func(ids = all_ids[1:limit])
  len_ids <- length(all_ids)
  num_pages <- floor(len_ids / limit)
  num_last_page <- len_ids - num_pages*limit
  page <- 1
  while(page != num_pages){
    print(glue("get_all_x_features: Processing page {page} of features"))
    new_features <- func(ids = all_ids[(page*limit + 1):(page*limit + limit)])
    features <- rbind(features,new_features)
    page = page + 1
  }
  print(glue("get_all_x_features: Processing page {page} of features"))
  new_features <- func(ids = all_ids[(num_pages*limit + 1):(num_pages*limit + num_last_page)])
  rbind(features,new_features)
}

#function POSTs tracks to a playlist that is owned by a certain spotify account
#very similar to previous pagination methods, except it adds tracks to a playlist
#instead of extracting info
add_all_tracks_to_playlist <- function(user_id, all_ids,auth_code,playlist_id,name){

  if(length(all_ids) <= 80){
    return(add_tracks_to_playlist(playlist_id = playlist_id, uris = all_ids))
  }
  print(glue("add_all_tracks_to_playlist: Adding songs 1-80 to playlist {name}"))
  add_tracks_to_playlist(playlist_id = playlist_id, 
                         uris = all_ids[1:80],
                         authorization = auth_code)
  len_ids <- length(all_ids)
  num_pages <- floor(len_ids / 80)
  num_last_page <- len_ids - num_pages*80
  page <- 1
  while(page != num_pages){
    print(glue("add_all_tracks_to_playlist: Adding songs {page*80}-{page*80 + 80} to playlist {name}"))
    add_tracks_to_playlist(playlist_id = playlist_id,
                           uris = all_ids[(page*80+ 1):(page*80 + 80)],
                           authorization = auth_code)
    page = page + 1
  }
  print(glue("add_all_tracks_to_playlist: Adding songs {num_pages*80}-{num_pages*80 + num_last_page} to playlist {name}"))
  add_tracks_to_playlist(playlist_id = playlist_id,
                         uris = all_ids[(num_pages*80 + 1):(num_pages*80 + num_last_page)],
                         authorization = auth_code)
}
  

```

```{r genres}
run_app <- function(auth_code){
  #handles ui logic for two apps; an app that combines playlists, and an 
  #app that visualizes and filters playlist data. 
  ui <- navbarPage(title = "Spotify Playlist Manager", 
                  tabPanel("Visualizations and Filtering",
                      sidebarLayout(
                        sidebarPanel = sidebarPanel(
                          selectInput("type","Type of Playlist",choices = c("Liked Songs","Playlist")),
                          conditionalPanel(
                            condition = "input.type == 'Playlist'",
                            helpText("A playlist ID can be found by getting the share link for a Spotify playlist and taking everything between playlist/ and the ? "),
                            textInput("id","Playlist ID", value = "2EgjrgDSeAF25yPq7No1mG"),
                          ),
                          checkboxInput("filter","Filter results?", value = FALSE),
                          conditionalPanel(
                            condition = "input.filter",
                            numericInput("followers_min","Minimum Followers",value = 0, min = 0),
                            numericInput("followers_max", "Maximum Followers",value = 9999999999, min = 0),
                            helpText("Enter all for genre to not filter by genre in output."),
                            textInput("genre","Genre"),
                            sliderInput("danceability","Danceability", min = 0, max = 1,value = c(0,1)),
                            sliderInput("energy","Energy", min = 0, max = 1, value = c(0,1)),
                            sliderInput("speechiness","Speechiness", min = 0, max = 1, value = c(0,1)),
                            sliderInput("valence","Valence",min = 0, max = 1, value = c(0,1)),
                            sliderInput("acousticness","Acousticness",min = 0, max = 1, value = c(0,1)),
                            sliderInput("tempo","Tempo (bpm)",min = 0, max = 500, value = c(0,500))
                          ),
                          checkboxInput("playlist","Create Playlist of Results?"),
                          conditionalPanel(
                            condition = "input.playlist",
                            textInput("user_name","Username"),
                            textInput("name","Name of Playlist")
                          ),
                          actionButton("run_viz","Get Songs")
                          ),
                        mainPanel = mainPanel(
                          tabsetPanel(
                            tabPanel("Songs", 
                                     DTOutput("songs")),
                            tabPanel("Spotify Variable Summary",
                                     plotOutput("graph_sum"),
                                     plotOutput("followers"),
                                     DTOutput("summary")),
                            tabPanel("Genre Summary",
                                     DTOutput("genre"))
                          )
                          
                        )
                      )
                    ),
                    tabPanel("Combine Two Playlists", #app for combining playlists
                             sidebarLayout(
                               sidebarPanel = sidebarPanel(
                                textInput("playlist1","Playlist 1 ID", value = "76ZauyHiAEI4ftqo6hfyt4"),
                                textInput("playlist2", "Playlist 2 ID", value = "3Nb7Xx6ohBfOHq0FVMMg07"),
                                textInput("userid","User_Name"),
                                textInput("new_name","New Playlist Name"),
                                actionButton("run_comb","Combine")),
                               mainPanel = textOutput("complete")
                             )
                            )
              )

  server <- function(input,output,session){
   
    observe({
      #following code produces the dataframe used for all of proceeding calculations
      #has genre information, spotify internal data (i.e. valence), date, 
      #number of followers for an artist, etc. 
      
      if(input$type == "Liked Songs")
        tracks <- get_all_saved_tracks(auth_code = auth_code) 
      else{
        tryCatch({ #used from midterm2 to check for errors,
        tracks <- get_all_playlist_tracks(id = input$id)
      },
      warning = function(warn){
          shinyalert(paste0(warn),
                     type = "warning",
                     size = "l")
      },
      error = function(err){
          shinyalert(paste0(paste0(err), "\n Check to make sure the Playlist ID was entered correctly."),
                     type = 'error',
                     size = "l")
        validate(need(exists("playlist"), message = FALSE))
        
        
      })
    }
      validate(need(exists("tracks"), message = FALSE)) #checks that result worked
      
      feats <- get_all_x_features(tracks$track.id, func = get_track_audio_features, limit = 100) %>% 
        select(-12,-(14:17))
      more_info_tracks <- full_join(tracks,feats, by = c("track.id" = "id")) %>% 
        unnest(track.artists)
      
      print("Processed all track features, grabbing genres of artists")
      artists <- get_all_x_features(more_info_tracks$id, func = get_artists, limit = 50) %>% 
        select(genres,followers.total,id)
      
      print("Finished processing genres of artists, cleaning up and combining")
      all_info_tracks <- unique(full_join(more_info_tracks,artists, by = "id"))
      print("Done!")
      
      all_info_tracks <- all_info_tracks %>% 
        mutate(added_at = as.Date(added_at),
               track.album.release_date = as.Date(track.album.release_date)
           ) %>% 
        rename(Artist = name, 
                 Track_Title = track.name,
                 Album = track.album.name,
                 Genre = genres,
                 Artist_Followers = followers.total)
     
      if(input$filter){ #if filtering is turned on, does the appropriate filtering
        all_info_tracks <- all_info_tracks %>% 
          filter(valence > input$valence[1] & valence < input$valence[2]) %>% 
          filter(energy > input$energy[1] & energy < input$energy[2]) %>% 
          filter(acousticness > input$acousticness[1] & acousticness < input$acousticness[2]) %>% 
          filter(speechiness > input$speechiness[1] & speechiness < input$speechiness[2]) %>% 
          filter(danceability > input$danceability[1] & danceability < input$danceability[2]) %>% 
          filter(tempo > input$tempo[1] & tempo < input$tempo[2]) %>% 
          filter(Artist_Followers > input$followers_min & Artist_Followers < input$followers_max) %>% 
          filter(str_detect(Genre, input$genre) | str_to_lower(input$genre) == "all")
        
      }
      if(nrow(all_info_tracks) == 0){ #if genre has no rows, tells the user before errors occur
        shinyalert(title = "Error",
                   "Result has no rows. Make your filter conditions more lenient.",
                   type = "error",
                   size = "l")
      }
      
      validate(need(nrow(all_info_tracks) > 0, message = FALSE))
      
      if(input$playlist){ #if user has asked for results to be put into playlist, checks to make sure it works
        tryCatch({ 
        playlist_filter <- create_playlist(user_id = input$user_name, 
                            name = input$name,
                            authorization = auth_code)
      },
      warning = function(warn){
          shinyalert(paste0(warn),
                     type = "warning",
                     size = "l")
      },
      error = function(err){
          shinyalert(paste0(paste0(err), "\n Make sure Username was entered correctly and playlist name is not empty."),
                     type = 'error',
                     size = "l")
      })
      validate(need(exists("playlist_filter"), message = FALSE))
      add_all_tracks_to_playlist(user_id = input$user_name, #adding tracks
                                 playlist_id = playlist_filter$id,
                                 all_ids = all_info_tracks$track.id,
                                 auth_code = auth_code,
                                 name = input$name)
    }
      
      output$songs <- renderDT({ #renders key aspects of playlist returned
        all_info_tracks %>% 
          select(Artist, Track_Title, Album, Genre, Artist_Followers) 
        })
      
      output$graph_sum <- renderPlot({ #CDF plot of internal variables
      all_info_tracks %>% 
        pivot_longer(cols = c(acousticness,speechiness,valence,danceability,energy),
                     names_to = "Spotify Internal Variables",
                     values_to = "spotify_val") %>% 
        ggplot(aes(x = spotify_val, color = `Spotify Internal Variables`)) + 
                 stat_ecdf() + 
        theme_bw() + 
        labs(x = "Values", 
             y = "Proportion of Songs Below Value",
             title = "Empirical CDFs of Internal Spotify Variables")
      })
      
      output$summary <- renderDT({ #summary stats
        all_info_tracks %>% 
            summarize(mean_danceability = mean(danceability),
            mean_speechiness = mean(speechiness),
            mean_energy = mean(energy),
            mean_valence = mean(valence),
            mean_tempo = mean(tempo),
            mean_acousticness = mean(acousticness),
            median_follower_count_of_artists = median(Artist_Followers)
            )
      })
      output$followers <- renderPlot({ #followers histogram
        ggoutlier_hist(all_info_tracks, 
                       var_name = "Artist_Followers", 
                       cut_off_ceiling = 
                         quantile(all_info_tracks$Artist_Followers,.9)) + 
          labs(x = "Followers", y = "Count", title = "Distribution of Artist Follows Below 90th Percentile") + 
          theme_bw()
      })
      output$genre = renderDT({ #summarizes genre of playlists
        all_info_tracks %>% 
          count(Genre) %>% 
          arrange(desc(n)) %>% 
          rename(`Number of Songs` = n)
          })
      
    }) %>% 
      bindEvent(input$run_viz)
  
    
    observe({ #for the 2nd app
      tryCatch({ #used from midterm2
        play1 <- get_all_playlist_tracks(id = input$playlist1) #reads in both playlists
        play2 <- get_all_playlist_tracks(id = input$playlist2)
        playlist <- create_playlist(user_id = input$userid, 
                            name = input$new_name,
                            authorization = auth_code)
      },
      warning = function(warn){
          shinyalert(paste0(warn),
                     type = "warning",
                     size = "l")
      },
      error = function(err){
          shinyalert(paste0(paste0(err), "\n Check to make sure the Playlist IDs/Username was entered correctly. Additionally, make sure New Playlist Name has a value."),
                     type = 'error',
                     size = "l")
      })
      
      #validates that playlist creation and grabbing each playlist was successful
      validate(need(exists("play1"), message = FALSE)) 
      validate(need(exists("play2"), message = FALSE))
      validate(need(exists("playlist"), message = FALSE))
      new_play <- rbind(play1,play2)
      #combines tracks 
      add_all_tracks_to_playlist(user_id = input$userid, 
                                 playlist_id = playlist$id,
                                 all_ids = new_play$track.id,
                                 auth_code = auth_code,
                                 name = input$new_name)
      #prints textoutput notifying user task succeeded. 
      output$complete <- renderPrint({"Successfully combined playlists"})
    }) %>% 
       bindEvent(input$run_comb)
  }
  
  observe({
    
  })
  
  shinyApp(ui,server)
}
```

```{r run app, eval = FALSE, message = FALSE}
run_app(auth_code)
```

## Discussion & Conclusions

While the app does everything I wanted to do for the most part, there is
still some room for improvement. In particular, there could be more 
visualization/analysis of the data itself (such as time-series analysis
with the Date column) but I wanted to keep the app focused on playlist
management rather than detailed analysis. Additionally, the genre filter could
be more robust, as it does not work well if you want to filter for several 
different genres together; however, you can filter by a superset
(e.g. genre = indie will return indie rock, indie pop, indietronic, etc.) so 
this is not as big a concern as it might appear at first. I also know that
I could have reduced the scope of my ask for OAuth 2.0, but I didn't feel like
playing around with what specific scopes I needed for my analysis. In the future
I plan to improve this, as I certainly need a lot less permissions than I ask 
for to make my app work. 

Overall, I feel satisfied with my app, although I could potentially make it
more pretty so it looks less like a ShinyApp. I do intend to return to this
later and improve it more, so any feedback would be appreciated! Thank you and 
all the TAs for such an interesting/useful course, its filled in a lot of gaps
in my knowledge and I enjoyed it a ton. Have a great summer!
